---
title: 'hetGP4cast: adventures in ecological forecasting'
tags:
- R
- forecasting
- ecology
- Gaussian process
- heteroscedasticity
date: "08 December 2023" 
output:
  html_document:
    df_print: paged
authors:
- name: Maike Holthuijzen
  orcid: "0000-0002-6870-3314"
  equal-contrib: yes
  affiliation: 1
- name: R. Quinn Thomas
  equal-contrib: yes
  affiliation: 1
- name: Cayelan C. Carey
  equal-contrib: yes
  affiliation: 1
- name: Robert M. Gramacy
  corresponding: yes
  affiliation: 1
bibliography: references.bib
aas-doi: "NA"
aas-journal: NA
affiliations:
- name: Virginia Tech, USA
  index: 1
---

# Summary
Climatological models are useful for forecasting ecological phenomena, as they can capture long-term trends and tendencies. Climatological models also provide important baseline forecasts for researchers developing and testing new ecological forecasting models; for example, climatological models can generate 'null model' forecasts that can be used to determine the relative improvement in skill of new models. Many ecological variables that are increasingly being forecasted are characterized by heteroscedasticity over time and other dimensions. However, climatological models typically do not account for input-dependent variability, limiting their use for ecological forecasting. The incorporation of a non-constant variance over at least one dimension would improve uncertainty quantification (UQ) of climatological forecast models and could dramatically improve the prediction accuracy and quality of UQ of baseline forecasts. Heteroscedastic Gaussian process models (hetGPs) are ideal for constructing climatological models; however, while hetGPs can be fitted with existing software packages (i.e. the `hetGP` package @binois2021hetgp in R), they are not as easily accessible to non-statisticians. In response to this need, we developed the `hetGP4cast` package to fit hetGPs to data targeted for environmental applications.

# Statement of need
Climatological forecasts are often used as baseline or reference models for ecological forecasting, but none currently account for heteroscedasticity. Typically, the mean and standard deviation of climatological forecasts used for ecological forecasting are calculated directly as statistics from long-term climatologies. Empirical means and standard deviations are suboptimal for forecasting, because unequal sample sizes used in calculations are ignored, making UQ less robust. There is a critical need for a more sophisticated statistical modeling approach that incorporates a non-constant variance. Such an approach would involve fitting a hetGP model to long-term climatological records. A hetGP climatological model would result in more accurate predictions and, importantly, much more reliable UQ, resulting in better quality of baseline forecasts used for ecological forecasting. The `hetGP4cast` package was specifically designed for this purpose. Unlike existing tools (i.e., the `hetGP` package by @binois2021hetgp), the `hetGP4cast` R package empowers ecological forecasters, especially those without an extensive background in Gaussian process modeling, to easily fit, generate predictions, and plot results from heteroscedastic Gaussian process models. By adopting `hetGP4cast`, ecologists and other non-statisticians can enhance the accuracy of their climatological models, resulting in more reliable predictions and, crucially, enhanced UQ. Looking ahead, the `hetGP4cast` package represents a valuable resource for the ecological forecasting challenge community. We believe that its intuitive interface and powerful capabilities will prove useful to ecologists aiming to construct robust climatological models. The package bridges the gap between statistical sophistication and accessibility, ensuring that even non-statisticians can contribute meaningfully to advancing cutting edge research in ecological forecasting.


## Examples
### Day of year models
In this section, we use the `hetGP4cast` package to fit hetGPs to sample data and construct 1 to 35-day ahead forecasts for a desired reference data. In this section, day of year (``DOY``, 1-366) will be the only input to the model. The `hetGP4cast` package is simple and includes 2 main functions: `fit_hetgp()` and `predict_hetgp()` for fitting hetGPs and predicting (forecasting). The parameters for `fit_hetgp()` are:


* `X:` model inputs (covariates). Can be either `DOY` or `c(DOY, <name of covariate>)`. `DOY` refers to using Julian day of year as a covariate. The additional covariate besides `DOY` must be numeric. Categorical variables are currently not supported.
* `Y:` response variable. Must be one of the unique values of the `variable` column in `df`.
* `site_id`: The focal site. Must be one or more of the unique values of the `site_id` columns of `df`.
* `df`: the input dataframe. Must be in standard format, see [here](https://projects.ecoforecast.org/tern4cast/instructions.html#target-data-calculation). The only other extra column that may be included must have a column name `depth` and should denote the depth of the temperature in meters. The `datetime` column must be formatted as YYYY-MM-DD, but it may either be of class `string, POSIXct` or `Date`.
* `covtype`: the type of covariance function used in fitting the hetGP. We recommend `Gaussian` (the default), but any one of `Gaussian`, `Matern3_2` or `Matern5_2` are acceptable (see the ``hetGP`` package by @binois2021hetgp for details). 
* `covar_levels` : the levels at which forecasts are to be evaluated if an additional covariate (other than DOY) in included. If not specified (`covar_levels`= `NULL`), and the number of levels are less than 10, those levels will be used; if there are more than 10 levels, 10 levels will be specified using the following quantiles: 0, 11, 22, 33, 44, 55, 66, 77, 88, 100. The user may also specify a numeric vector of levels.
* `silent`: if `TRUE` additional output will be printed to the console.

`fit_hetgp()` will return errors if `df` is not of standard format (has extra or missing columns or if column classes are incorrect). In addition NAs for inputs `X` and the response variable `Y` are not allowed. However, the function `fit_hetgp()` will return informative errors.

The sample dataset used here contains lake temperatures for several reservoirs located in Southwest Virginia, USA.
```{r}
library(hetGP4cast)

# read data
lakedat = readr::read_csv("https://zenodo.org/records/10114876/files/lakedat.csv", show_col_types = FALSE)
lakedat = lakedat[, 2:ncol(lakedat)]

# extract only 00UTC
lakedat$hour = format(lakedat$datetime, format = "%H:%M:%S")

lakedat = lakedat[lakedat$hour == "00:00:00", ]
lakedat$hour = NULL

# extract only 1m depth observations; get rid of depth column; eliminate NAs
lakedat = lakedat[complete.cases(lakedat), ]
lakedat1 = lakedat[lakedat$depth == .1, ]
lakedat1$depth = NULL

# inspect resulting data.frame
head(lakedat1)
```

Here we fit a climatological model for the `SUGG` site.
```{r}
model1 = fit_hetgp(X = "DOY", Y = "temperature",
                   site_id = "SUGG", df = lakedat1)
```


`fit_hetgp()` returns a list with the following items:
* hetGP fit object containing information about the model fit
* `df`: the original input dataframe 
* `include_covar`: boolean to denote whether or not depth was used as a covariate
* `Yname`: values of the response variable
* `covar_levels`: levels of the covariate at which predictions (forecasts) are to be made
* `Y_resp`: a vector containing the values of the response variable

To make forecasts, we need only to pass the return value of `fit_hetgp()` to `predict_hetgp()`. `predict_hetgp()`'s parameters include:
* `hetGP_fit:` hetGP object fit from `fit_hetgp()`
* `save_covmat:` boolean, denotes whether the predictive covariance matrix should be saved (default is \texttt{FALSE})
* `reference_datetime`: the reference date for which forecasts are desired. `reference_datetime` can be either a string or of class `Date` or `POSIXct` and must be formatted as `YYYY-MM-DD`.
* `max_horizon`: the maximum forecast horizon in days (default is 35).
* `PI`: value in [0,1) to denote level of prediction intervals. We recommend `PI` = 0.90 or 0.95 (default is 0.90).

```{r}
preds1 = predict_hetgp(het_gp_object = model1, reference_datetime = "2023-09-01")
```

`predict_hetgp()` returns a list containing the following items:

* `pred_df`: a `data.frame` in standard format containing `max-horizon`-day ahead forecasts (predictions from the fitted hetGP model) for the desired `reference_datetime`
* `covmat`: the predictive covariance matrix (if `save_covmat` = TRUE; otherwise `covmat=NULL`. Note that the nugget term is included on the diagonal of `covmat`, resulting in predictive (not confidence) intervals.
* `df` the original input `data.frame` for `fit_hetgp()`
* `preds4plotting`: a `data.frame` formatted for easy plotting; includes columns `Mean` (predictive mean), `sd` (standard deviation plus nugget term), `DOY`, `Lower` and `Upper` (lower and upper bounds for prediction intervals).
* `include_covar`: boolean to denote whether or not another covariate in addition to `DOY` was included
* `Yname`: name of the response variable `Y`
* `pred_width`: width of prediction intervals (value in `[0,1)`)
* `covar_levels`: levels of the covariate at which predictions are to be made
* `covar_name`: name of the additional covariate, if applicable


Now, we can quickly plot the forecast and prediction intervals using `plot_hetGPpreds()`. 
```{r, fig.dim=c(6,6)}
plot_hetGPpreds(predObject = preds1)
```

## Climatological models with 2 covariates
Models with 2 covariates can take some time to fit. To keep computation time down, we only use depths 1 and 2 for this example. If there are not enough replicates over `depth` and `DOY`, parameter estimation is not possible, and `fit_hetGP()` will return an error (you can check the number of replicates by running `table(df$depth, df$DOY)`, where `df` is the input dataframe, after converting the `datetime` column to `DOY` (Julian day).


We can proceed exactly as before, but this time we include `depth` as a covariate. We do not recommend fitting models to datasets of more than 5,000 observations. When the input `df` contains the extra input variable `depth`, the user may elect to instead run a 1-input variable using only `DOY` as the input. However, we do not recommend doing so, as the model will pool observations from all depths.

```{r}
# extract only depths 1 and 2
lakedat2 = lakedat[lakedat$depth %in% c(0.1, 1.1), ]
# takes a little bit to run
model_depth = fit_hetgp(X = c("DOY","depth"), Y = "temperature",
                   site_id = "SUGG", df = lakedat2)
```


We can use the `predict_hetgp()` function as before, but we specify `depths` to be `1:2`.
```{r}
modeld_preds = predict_hetgp(het_gp_object = model_depth, reference_datetime = "2023-09-01")
```

As before, the results can be plotted using `plot()`.
```{r, fig.dim=c(8,6)}
plot_hetGPpreds(predObject = modeld_preds)
```

## Advanced 

To construct realizations from the posterior predictive distribution, only the predictive mean and covariance matrix are necessary. Suppose we have some data $D$ composed of $(x,y)$ pairs, and we are interested in determining which function $f(x)$ generated the data $D$. If we assign a GP prior to $f$, we have $f \sim GP$. Given this setup, the posterior distribution $f(x)|D$ is generally of interest. Since we are usually interested in predictions at new values of $x$, $x*$, we usually want the posterior predictive distribution. Fortunately the posterior predictive distribution is also multivariate normal. That is, for several new $x$ locations $\mathcal{X}$ we are interested in $Y(\mathcal{X}) \mid D$: $Y(\mathcal{X}) \mid D \sim \mathcal{N}(\mu(\mathcal{X}), \Sigma(\mathcal{X}))$. Both $\mu(\mathcal{X})$ and $\Sigma(\mathcal{X})$ can be obtained from `predict_hetgp()`. Note that simulations performed in this manner will give rise to predictive (not confidence) intervals.

```{r}
preds2 = predict_hetgp(het_gp_object = model1, reference_datetime = "2023-09-01",
                       save_covmat = TRUE)
dim(preds2$covmat)
```

The `hetGP4cast` package includes a function to generate realizations from the posterior predictive distribution. We need only to pass the prediction object from `predict_hetGP()` (with `save_covmat=TRUE`) to the function `make_realizations()`. The number of realizations desired is specified with `nreals`.
```{r}
mySims = make_realizations(predObject = preds2, nreals = 200)
```

The function returns a list containing a data.frame in standard format including the realizations as well as a matrix containing the realizations.

```{r}
head(mySims$pred_df_reals)
```


# Acknowledgements

We acknowledge the Ecological Forecasting Project team (especially Mary Lofton, Freya Olsson, and Austin Delany) for helpful comments and feedback. We acknowledge funding from the U.S. National Science Foundation grants 1933016, 2327030, and 2318861.

# References
@article{binois2021hetgp,
  title={hetgp: Heteroskedastic Gaussian process modeling and sequential design in R},
  author={Binois, Micka{\"e}l and Gramacy, Robert B},
  year={2021}
